---
title: "Introduction to Quarto"
format: html
bibliography: references.bib
---

# Introduction

<!-- -->

## About the data

-   Water sample data from Arctic Ocean
-   Data downloaded from the Arctic Data Center: [Tweedie ADC data](http://doi.org/10.18739/A25T3FZ8X)

## Setup

```{r}
#| warning: false
library(readr)
library(here)
```

## Read in data

```{r}
#| warning: false
bg_chem <- read_csv(here("data/BGchem2008data.csv"))
```


# Analysis

## a. Calculate summary statistics
<!-- -->
```{r}
#| eval: false
#| echo: false

# Print column names
colnames(bg_chem)

#show the data structure
str(bg_chem)

#show first few lines
head(bg_chem)

#summary statistics of each column
summary(bg_chem)

#print unique values in a specific column
unique(bg_chem$Date)
```

## b. Calculate mean Redfield ratio
```{r}
# calculate means
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```


## c. Plot Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
```

Here is the ratio we calculated: `r ratio`  
Here is the ratio we calculated: `r round(ratio)`

# Conclusion






Markdown formatting

# This is a large title

## this is a smaller title

##### this is a tiny title

*text italics* **bold**

## station data

1.  here we will load our .csv files
2.  here is where we define the datetime format and force the time zone to AKST
3.  The third step is to look for missing [@oke2020] data.

superscript^2^

subscript~2~

~~strikethrough~~

| A   | B   | C     |
|-----|-----|-------|
| 1   | x   | test  |
| 2   | y   | test1 |
| 3   | z   | test2 |
