---
title: "Tidying and Wrangling"
format: html
---

```{r}
#| message: false
library(dplyr)
library(tidyr)
library(readr)
```

# Data Sources

## Read in our data
```{r}
#| message: false
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

```

## Explore the data
```{r}
colnames(catch_original)
head(catch_original)
summary(catch_original)
unique(catch_original$Region)
# control shift M is %>% !!!! wow
```

# Cleaning and reshaping 

## Using the select() function to select columns in our dataset

```{r}
#select() to keep the columns we want
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)
# OR we could just drop columns we don't want
catch_data<- catch_original %>% 
  select(-All,-notesRegCode)

```

##check the column types using glimpse()
```{r}
glimpse(catch_data)
# we see chinook is a character
```


## changing column contents with mutate()
```{r}
#change chinook column to numeric
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))


```
```{r}
#there were some na characters show them
is.na(catch_clean$Chinook)
#get the location of the na value, if found call it an i
i <- which(is.na(catch_clean$Chinook))
#Look a the specific place where it is occurring (row) 
catch_data[i,]
#this was a pdf originally and was converted to a table so the I showed up.

#use mutate to fix that: if chinoook = I then convert to a 1 or just use what is already in chinook and then save as numeric (instead of character)
catch_clean<-catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I", 
                           true="1", 
                           false = Chinook),
         Chinook = as.numeric(Chinook))
head(catch_clean)
```
## Changing file shape using pivot_longer() and pivot_wider()
```{r}
#decrease the columns of all the types of salmon and pivot them into rows (tidy format) new name will be species
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to ="catch")

head(catch_long)
#now change it back to the wide format, move the species back into their own columns
catch_wide<- catch_long %>% 
  pivot_wider(names_from = species, 
              values_from = catch)
head(catch_wide)
```
## rename the columns with rename()
```{r}
# the column with catch is in the thousands
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)
head(catch_long)

```

```{r}

#add a new column with the actual catch value (by multiplying by a thousand) and drop the catch thousands from the dataset
catch_long <- catch_long %>% 
  mutate(catch=catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)


```
##summary statistics using group_by and summarize()
initial data - split by region - apply - to get mean catch per region - lastly combine into summary statistics
```{r}
#get average catch by region
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch))
head(mean_region)

#number of rows for each group using a function n() to get the counts
n_region <- catch_clean %>% 
  group_by(Region) %>% 
  summarise(n = n())
head(n_region)

```
#Filtering rows using filter()
```{r}
#filter by one specific region SSE, return all the rows where region is SSE
sse_catch<-catch_long %>% 
  filter(Region == "SSE")
head(sse_catch)

#filter catch_long where the catch column is over 1 million fish
catch_million <- catch_long %>% 
  filter(catch > 1000000)
head(catch_million)

#filter the rows where the region is SSE and chinook
chinook_sse <- catch_long %>% 
  filter(Region == "SSE" & species == "Chinook")

head(chinook_sse)

```

##Sorting data with arrange() to look at data in a specific order

```{r}
#look at the mean catch by region and arrange by mean catch
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(mean_catch)
head(mean_region)
#show mean catch in  descending order desc()
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))
head(mean_region)

```
## combinging datasets using left join
```{r}
#get the region file and add the management area to the catch long file but call it catch_joined
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  select(code,mgmtArea)
head(region_defs)
catch_joined<-left_join(catch_long, region_defs, by=c("Region" = "code"))
head(catch_joined)
```
```{r}
#or do it this way using rename
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  select(code,mgmtArea) %>% 
  rename(Region = code, Region_Name = mgmtArea)

catch_joined <- left_join(catch_long, region_defs, by = c("Region"))
head(catch_joined)
```
## split and combine values in columns

```{r}
#create site names/nuimbers
sites_df <- data.frame(site = c("HAW-101", "HAW-103","OAH-320"))
head(sites_df)
# split site names into 2 columns using separate()
split_df <- sites_df %>% 
  separate(site, c("island","site_number"), sep = "-")
head(split_df)
```

```{r}
dates_df<-data.frame(
  year = c("1930", "1930", "1932"),
  month= c("12", "12", "12"),
  day = c("14", "15", "16")
)
head(dates_df)
#unite the 3 columns into one column usiing a separator and call the column "date"
dates_unite <- dates_df %>% 
  unite(col=date, year, month, day, sep = "-")
head(dates_unite)


```
```{r}
#this could all be done in 1 step
mean_region <- catch_original %>% 
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year),
               names_to = "species",
               values_to = "catch") %>% 
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)

```
#save
```{r}
library(here)
write_csv(mean_region, here("data/mean_catch_by_region.csv"))
```

